# EnSDD: Quick Start

Do you like to live in the fast lane? Do you sigh every time you see a
long vignette? Do you wish there was a TLDR on everything?

I understand.

Welcome to the EnSDD quick start!

In this vignette, we reproduce the results shown in fig 4 in the article.

# Introduction

Intratumoral heterogeneity in cancer poses challenges for formulating effective treatments and induces poor survival outcomes. Spatial transcriptomics is a valuable tool for dissecting and characterizing the heterogeneity, offering insights into both the complexity of tumor and the interplay between tumor and immune responses. To illustrate the
generalization power on cancer tissue, we test EnSDD on the human breast cancer dataset.

This experment includes several steps: (1) running ensemble-based learning methods for the identification of spatial domains; (2) Based on the spatial domains information, identifying the domain-specific SVGs; (3) inferring the cell type spatial distribution.

# Loading and Pre-processing

First, let’s load the package and the data. We focus on human breast
cancer data from 10x Visium technology. The data includes spatial gene expression, spatial meta information and morphology image. We provide the data at the \[Zenodo data repository\](<https://doi.org/10.5281/zenodo.4739739>). The data contains 3798 spots and 36601 genes.                             


```{r}
suppressMessages(library(EnSDD))
#download the data and load the data
python_env <- "~/.conda/envs/EnSDD/bin/python"
counts_path = "/home/vision/Documents/EnSDD/data/counts_matrix.txt"
loc_path = "/home/vision/Documents/EnSDD/data/meta_data.txt"
img_path = "/home/vision/Documents/EnSDD/data/V1_Breast_Cancer_Block_A_Section_1_image.tif"
reticulate::use_python(python_env, require = T)
reticulate::py_config()
Seurat.data <- data_process(counts_path, loc_path, img_path, n_HVG = 2000, n_PCA = 30)

```

# Spatial domain detection

For the spatial domain detection, EnSDD firstly run six base SDD methods
and obtain multiple clustering results. For each clustering vector, a binary similarity matrix is then constructed based on the corresponding spot labels: if two spots belong to the same cluster, their similarity is set to 1; otherwise, it is 0. Utilizing the binary similarity matrices generated by the base SDD methods, EnSDD introduces an optimization model to obtain a consensus similarity matrix of spots. This model aims to simultaneously derive the fully spatial similarity matrix between spots and the adaptive weights assigned to each base method. These learned weights are proved to be highly correlated with the performance of base methods, and can be further used as an evaluation metric for base spatial clustering methods. Finally, EnSDD introduces the Louvain algorithm with adaptive resolution for the identification of spatial domains based on the consensus similarity matrix of spots.


```{r}
## setting the number of clustering equal to the annotation provided by SEDR
label_setting <- 20
# BayesSpace will take long time to run, don't run setting BayesSpace = FALSE
# The user could download the results of base methods that provided in the ./data/res.clustering.20.RData
# For reproducing the results, we load the .RData
# The user enable to run the run_individual_clustering function for the applying own data

# res <- run_individual_cluster(counts_path, loc_path, img_path, python_env, n_HVG = 2000, n_PCA = 30,number.setting.by.user = label_setting ,saving_results = TRUE)

load("../data/res.clustering.20.RData")

# getting the consensus matrix based on multiple binary similarity matrix
res.ensemble <- solve_ensemble(res$clustering.mat, lambda = NULL, prob.quantile = 0.5,niter = 100, epsilon = 1e-5, ncors.used = 24)
# using louvain algorithm with adaptive resolution for the detection of spatial domains
res.ensemble.clustering.louvain <- Louvain_clustering(Seurat.data, similarity_mat = res.ensemble$H,setting_k = label_setting, max.res = 5, resolution_louvain = NULL, increase = 0.05)

### ARI and Pur values
meta_data <- Seurat.data@meta.data
res.vec <- res$clustering.vec
res.vec.df <- as.data.frame(res.vec)
res.vec.df$EnSDD <- res.ensemble.clustering.louvain[rownames(meta_data)]
meta_data_all <- cbind(meta_data, res.vec.df)

methods_names <- c(names(res$clustering.vec), "EnSDD")
ARI_res <- c()
for(i in 1:length(methods_names)){
  ari_res <- mclust::adjustedRandIndex(meta_data_all$old_fine_annot_type,    meta_data_all[[methods_names[i]]])
  ARI_res <- c(ARI_res, ari_res)
}
names(ARI_res) <- methods_names
df.bar <- data.frame(Method = methods_names, ARI = ARI_res)
library(ggplot2)
title_sample <- "Human breast cancer"
error_bar = ggplot(df.bar, aes(x = Method , y = ARI,
                                  fill = Method))+
    geom_bar(stat="identity")+
    geom_text(aes(label = round(ARI, 4)), vjust=0)+
    theme_classic()+
    labs(y="ARI", x=" ", title = title_sample)+
    theme(
      axis.text.x.bottom = element_text(size = 14,hjust = 1,angle =45),
      axis.text.y.left = element_text(size = 12),
      axis.title.x = element_blank(),
      axis.title.y = element_text(size = 14),
      axis.ticks.x = element_blank(),
      plot.title = element_text( size=14,hjust = 0.5),
      legend.position = "none",
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_blank(),
      axis.line = element_line(colour = "black")
    )

library(reticulate)
reticulate::source_python(system.file("python", "Visualization_HBC_clustering.py", package = "EnSDD"))
write.table(meta_data_all, file = "res.clustering.txt", sep = "\t")

visualization_clustering(counts_path, "res.clustering.txt", img_path, cluster_num = as.integer(20), spot_diameter_fullres = 150, spot_size = 1.5)
unlike('res.clustering.txt')
# visualization_clustering(counts_path, "~/data/clustering_res.20.txt", img_path, cluster_num = as.integer(20), spot_diameter_fullres = 150, spot_size = 1.5)
```

# Identifying domain-specific spatially variable genes

EnSDD provides two ways for the identifying the domain-specific SVGs: Wilcoxon test and Local Getis and Ord’s Gi. For the human breast cancer, we define the domain 10 and 12 domain-specific SVGs.

```{r}
Seurat.data@meta.data <- meta_data_all
index.domain10_12 = Seurat.data$EnSDD == "10" | Seurat.data$EnSDD == "12"
# 
counts_ST <- as.matrix(Seurat.data@assays$SCT@counts)[, index.domain10_12]
meta_data_de <- Seurat.data@meta.data[index.domain10_12,]

suppressWarnings({Seurat.ST.DE <- Seurat::CreateSeuratObject(counts = counts_ST, assay = "Spatial", meta.data = meta_data_de)})
Seurat.ST.DE <- Seurat::SCTransform(Seurat.ST.DE, assay = "Spatial", verbose = FALSE)
  
### select HVG genes
Seurat.ST.DE <- Seurat::FindVariableFeatures(Seurat.ST.DE, nfeatures = 2000, verbose = FALSE)
### PCA 
Seurat.ST.DE <- Seurat::ScaleData(Seurat.ST.DE, verbose = FALSE)
Seurat.ST.DE <- Seurat::RunPCA(Seurat.ST.DE, npcs = 20, verbose = FALSE)
### UMAP of subtype
# Seurat.ST.DE <- Seurat::RunUMAP(Seurat.ST.DE, reduction = "pca", dims = 1:20)
# p1 <- Seurat::DimPlot(Seurat.ST.DE, reduction = "umap", group.by = "EnSDD")


labels_de <- Seurat.ST.DE@meta.data$EnSDD
res.temp <- Seurat::FindMarkers(Seurat.ST.DE, ident.1 = levels(as.factor(labels_de))[1], ident.2 = levels(as.factor(labels_de))[2], group.by = "EnSDD",test.use = "wilcox", logfc.threshold = 0.25)
index.de <- res.temp$p_val <= 0.05 & res.temp$p_val_adj <= 0.1
res.de <- data.frame(gene_names = toupper(row.names(res.temp)[index.de]),
                     log2FC = res.temp$avg_log2FC[index.de],
                     pvalue = res.temp$p_val[index.de],
                     FDR = res.temp$p_val_adj[index.de])

### loc_G ###
### HEBP1, FAM234b, CSTA, TMSB10
Loc_G_mat <- matrix(0, nrow = ncol(Seurat.ST.DE), ncol = 4)
rownames(Loc_G_mat) <- colnames(Seurat.ST.DE)
colnames(Loc_G_mat) <- c("HEBP1", "FAM234B", "CSTA", "TMSB10")

for (i in 1:ncol(Loc_G_mat)) {
  Loc_G_mat[,i] <- EnSDD::LocalG_spa(Seurat_data = Seurat.ST.DE, gene_name = colnames(Loc_G_mat)[i], k = 4)
}
# write.table(Loc_G_mat, file = "Loc_G_mat.txt", sep = "\t")

Loc_G_df <- data.frame(locG = as.vector(Loc_G_mat), 
                       gene = rep(colnames(Loc_G_mat), each = nrow(Loc_G_mat)),label = as.character(rep(Seurat.ST.DE$EnSDD, time = ncol(Loc_G_mat))))
library(ggplot2)
ggplot_LocG <- ggplot(Loc_G_df, aes(x = gene, y = locG, fill = label)) + 
  geom_boxplot() +
  theme_classic()+
  labs(y="LGO's Gi", title = "Local Getis and Ord's Gi")+
  theme(
    axis.text.x.bottom = element_text(size = 14,hjust = 1,angle =45), 
    axis.text.y.left = element_text(size = 14),
    axis.title.x = element_blank(), 
    axis.title.y = element_blank(),#element_text(size = 14),
    plot.title = element_text( size=14,hjust = 0.5),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 14),
    # legend.position = "none",
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(colour = "black")
  ) 

ggplot_LocG
# ggsave(ggplot_LocG, filename = "LocG_box.pdf", width = 6, height = 6)


#### enrichment analysis
library(biomaRt)
library(clusterProfiler)
library(dplyr)
library(org.Hs.eg.db)
library(ggplot2)

DE.genes <- toupper(res.de$gene_names)
hmark = biomaRt::useEnsembl(biomart ="ensembl", dataset = "hsapiens_gene_ensembl", mirror = "asia")
tmp <- biomaRt::getBM(attributes = c("hgnc_symbol", "entrezgene_id"), filters = "hgnc_symbol",values = DE.genes, mart = hmark, uniqueRows = TRUE)

#### For pathway analysis 
tmp <- dplyr::filter(tmp, entrezgene_id != "NA")
tmp.unique <- tmp[!duplicated(tmp$hgnc_symbol),]
rownames(res.de) <- res.de$gene_names
res.de <- res.de[tmp.unique$hgnc_symbol,]
res.de$gene_names <- tmp.unique$entrezgene_id

genelist <- res.de$log2FC
names(genelist) <- res.de$gene_names
genelist = sort(genelist, decreasing = TRUE)

ego2 <- clusterProfiler::gseGO(geneList = genelist, 
                               ont = "BP",
                               OrgDb = org.Hs.eg.db,
                               pvalueCutoff = 0.05,
                               pAdjustMethod = "BH")

# ego2 <- clusterProfiler::gseKEGG(geneList = genelist, # ordered named vector of fold changes (Entrez IDs are the associated names)
#                                  organism = "hsa", # supported organisms listed below
#                                  nPerm = 1000, # default number permutations
#                                  verbose = TRUE)
#head(ego2, 3) 
library(enrichplot)
library(dplyr)


# %>%  barplot(x = "qscore") + xlab(expression(-log[10](FDR)))
### show header 9 items 
all_items <- ego2@result[order(ego2@result$p.adjust, decreasing = FALSE), ]
ggplot2_enrich = all_items[1:10,]
ggplot2_enrich$log_q.value = -log10(ggplot2_enrich$p.adjust)
tt = expression(-log[10](FDR))

ggplot2_enrich$Description <- factor(ggplot2_enrich$Description, 
                                     levels = ggplot2_enrich$Description[order(ggplot2_enrich$log_q.value, decreasing = FALSE)])

ggplot2_enrich$ID <- factor(ggplot2_enrich$ID, 
                                     levels = ggplot2_enrich$ID[order(ggplot2_enrich$log_q.value, decreasing = FALSE)])

plot_enrich <- ggplot(data = ggplot2_enrich, aes(x = log_q.value, y = ID)) +
  geom_bar(stat="identity", width = 0.7) +
  theme_classic() + 
  labs(x = tt, y = '') +
  theme(
    axis.text.x.bottom = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    # axis.title.x = element_text(size = 10),
    # axis.title.y = element_text(size = 10),
    # axis.ticks.x = element_blank(),
    plot.title = element_text( size=14,hjust = 0.5),
    legend.position = "none",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(colour = "black")
)
plot_enrich
```

# Inferring the cell type distribution

EnSDD employs RCTD to infer the cell type composition for SRT. Given the spatial domains and the cell type compositions in spots, EnSDD explores the differences of cell type abundances between spatial domains, which further dissects the cell type heterogeneity of the tissue. We use the run.RCTD function in the spacexr R package for the inferring the cell
type abundances for the SRT data. We download the reference of human breast cancer from DISCO (<https://www.immunesinglecell.org/>).
|  Cell type   | Number of cells |
|:------------:|:---------------:|
|   B cells    |       3206       |
|   CAFs       |       6573       |
|   Cancer epithelial    |       24489       |
|   Cycling T cells      |       1528       |
|   Cycling myeloid    |       463       |
|   DCs       |       955       |
|   Endothelial    |       7605       |
|   Macrophage       |       5929       |
|   Monocyte   |       2328       |
|   NK cells       |       1864       |
|   NKT cells    |       1122       |
|   Normal epithelial      |       4355       |
|   Plasmablasts    |       3524       |
|   PVL       |       5423       |
|   T cells CD4+    |       19231       |
|   T cells CD4+       |       11487       |

```{r}
load("../data/breastdata_rctd.RData")
sc_exp = breastdata_rctd$sc_exp
sc_label = breastdata_rctd$sc_lable
spot_exp = breastdata_rctd$sp_exp
spot_loc = breastdata_rctd$sp_loc

database = EnSDD::data_process_rctd(sc_exp, sc_label, spot_exp, spot_loc)
### rename the cell type for RCTD
# database$sc_label <- gsub("/", "_", database$sc_label)

rctd_res <- EnSDD::RCTD_run(database, CELL_MIN_INSTANCE = 20)

###### visualization of pie 
library(RColorBrewer)
library(ggplot2)
library(scatterpie)
library(cowplot)
library(gridExtra)
library(viridis)
library(rstatix)
library(ggpubr)

# load("breastdata_rctd.RData")
# load("rctd_HBC.RData")

# EnDecon_res = Results.dec.mouse[[1]]
EnDecon = rctd_res
celltype = colnames(EnDecon)
# celltype

spot_loc = breastdata_rctd$sp_loc
colnames(spot_loc) <- c("x", "y", "region")
spot_loc$region <- paste0("domain", spot_loc$region)
# data("breast.spot.annotation")
# spot_loc$region = breast.spot.annotation
spot_exp = breastdata_rctd$sp_exp

#setting the color
qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector  <-  unlist(mapply(brewer.pal, qual_col_pals$maxcolors, 
                              rownames(qual_col_pals)))
col_low <- "green"; col_high <- "red"

piedata_temp <- data.frame(x = spot_loc$x, y= spot_loc$y,
                           group =factor(1:length(spot_loc$x)))

pos_pie = as.data.frame(EnDecon)

piedata = cbind(piedata_temp,pos_pie)
##########
col_df <- data.frame(cell_types = celltype,
                     col_vector = col_vector[1:length(celltype)])

p1 = ggplot() + geom_scatterpie(aes(x=x, y=y, group=group),
                                cols= names(piedata)[-1:-3],
                                data = piedata,color=NA,
                                pie_scale = 1.3)+
  coord_equal() +theme_bw()+
  labs(y=" ",x=" ",title =" ")+
  theme(axis.ticks.x  = element_blank(), 
        axis.ticks.y  = element_blank(), 
        axis.text.x.bottom =element_blank(), 
        axis.text.y.left = element_blank(), 
        legend.text = element_text(size = 12,angle= 0.5), 
        legend.title = element_blank(), 
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        axis.line = element_blank())

p2 = p1 + scale_fill_manual(values = col_df$col_vector,
                            breaks = col_df$cell_types)
# save picture
#ggsave(p2 , file='pirchart.pdf',width = 7, height = 5) 
print(p2)

### cell type abundance in different regions
anno_type = names(table(spot_loc[,3]))
# ind_caner = which(spot_loc[,3] == anno_type[3])
domain10 = which(spot_loc[,3] == anno_type[1])
domain12 = which(spot_loc[,3] == anno_type[2]) 
#wilcox.test()
EnDecon_plot <- matrix(0, nrow = nrow(EnDecon), ncol = 4)
rownames(EnDecon_plot) <- rownames(EnDecon)
colnames(EnDecon_plot) <- c("Cancer Epithelial", "Macrophage", "DCs", "T cells CD4+")

EnDecon_plot[,1] <- EnDecon[,"Cancer Epithelial"]
EnDecon_plot[,2] <- EnDecon[,"Macrophage"]
EnDecon_plot[,3] <- EnDecon[,"DCs"]
# EnDecon_plot[,4] <- EnDecon[,"Cycling T-cells"]
# EnDecon_plot[,5] <- EnDecon[,"NKT cells"]
# EnDecon_plot[,6] <- EnDecon[,"NK cells"]
EnDecon_plot[,4] <- EnDecon[,"T cells CD4+"]
# EnDecon_plot[,8] <- EnDecon[,"T cells CD8+"]
# EnDecon_plot[,3] <- rowSums(EnDecon[, c("Macrophage", "DCs", "Monocyte")])
# EnDecon_plot[,4] <- EnDecon[,"NK cells"]
celltype <- colnames(EnDecon_plot)
pval = matrix(NA, length(celltype), 1)
rownames(pval) = celltype
colnames(pval) = c("comparing")
plotreg = list()
for( k in 1: length(celltype)){
  # print(k)
  cell_type_prop = EnDecon[,celltype[k]]
  domain10_prop  = cell_type_prop[domain10]
  domain12_prop = cell_type_prop[domain12]
  # immu_prop = cell_type_prop[ind_immu ]
  pval[k,1] = wilcox.test( domain10_prop ,domain12_prop)$p.value
  # pval[k,2] = wilcox.test( cancer_prop ,immu_prop )$p.value
  # pval[k,3] = wilcox.test( conn_prop ,immu_prop )$p.value
  region = c("domain10","domain12")
  reg_lable =  rep(region,c(length(domain10_prop),length(domain12_prop)))
  data_regcan = data.frame(porp = c(domain10_prop, domain12_prop),
                           type = factor(reg_lable, levels = region))
  # Wilcox test
  stat.test <- data_regcan %>%
    wilcox_test(porp ~ type) %>%
    add_significance()
  bxp <- ggboxplot(data_regcan, x = "type", y = "porp", fill = "type", 
                   palette = c("#d9b1f0", "#FF99CC"))+
    theme_classic()+
    labs(y="",x=" ",title = celltype[k])+
    theme(
      axis.text.x.bottom = element_text(size = 9,hjust = 0.5,angle = 45), 
      axis.text.y.left = element_text(size = 9),
      axis.title.x = element_text(size = 9,hjust = 0.5), 
      axis.title.y = element_blank(),#element_text(size = 14),
      plot.title = element_text( size=9,hjust = 0.5),
      legend.title = element_blank(), 
      legend.position = "none",
      panel.grid.major = element_blank(), 
      panel.grid.minor = element_blank(),
      panel.border = element_blank(),
      axis.line = element_line(colour = "black")
    ) + scale_x_discrete(
      breaks = c("domain10", "domain12"),
      label = c("domain10", "domain12")
    )
  # Box plot
  stat.test <- stat.test %>% add_xy_position(x = "type")
  plotreg [[k]]  = bxp + stat_pvalue_manual(stat.test, label = "p", 
                                            tip.length = 0.01)
}

print(plot_grid(plotlist = plotreg,nrow = 1))
```


